Changes in Faculty Composition and Stagnating State Support for Higher Education

### 2022-11-20 text/paper.tex
Updated draft.

\usepackage{subcaption}                                                    % Sub figures.

@@

\def\equationautorefname~#1\null{Equation~(#1)\null}                       % Fix autoref for quations

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy\thanks{
    Cornell University, Department of Economics.}
    \thanks{I thank Professor Evan Riehl for guidance on this project, Michael Lovenheim, Ronald Ehrenberg, Douglas Miller, and Francine Blau for various comments on the project's direction,
    as well as seminar participants at Cornell University (Autumn 202) for helpful
    discussion.
    This project's Github repository, with materials and (publicly available) data for replication, is available at 
    \url{https://github.com/shoganhennessy/finances-faculty-composition}.
    Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.}}
% Title
\title{Stagnating State Support for Higher Education and Its Effect on Faculty Composition}
\date{\today}
\rhead{\today}
\lhead{State Support for Higher Education and Faculty Composition}
% Begin
\begin{document}
\clearpage \maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
\noindent
State support for higher education has stagnated on a per student level the last thirty years, while public universities have reduced the number of tenure-track or with tenured professors per student, and increased their employment of contingent lecturers.
This analysis relates these changes in the faculty composition at public universities to the fall in per student university revenues from state appropriations.
A shift-share instrumental variables approaches addresses endogeneity in decisions for state appropriations to higher education, by exploiting the yearly change in state support with an institution's reliance on that support.
A decrease in state funding of 10\%, via a shock to state appropriations, decreases the number of assistant professors per student at a public university by 1.4\% and full professors by 1.2\%, yet increases the number of lecturers per student by 4.4\%.
Analysis of all the professors hired at Illinois public universities 2011-2021 shows that incumbent professors are not affected by the changes in state support, implying that these changes in faculty composition arose by lower rate of hiring of tenure-track and tenured professors at public universities.
These results show one of the long-term effects of stagnating state support for higher education, and raises questions about the direction that public education heads as these financial headwinds show no sign of dissipating.

\vspace{0.75cm}
\noindent\textbf{Keywords:}
State and Local Budget and Expenditures,
Higher Education,
Public Sector Labor Markets

\vspace{0.5cm}
\noindent\textbf{JEL Codes:} H72, I23, J45
\end{abstract}
\newpage
\setcounter{page}{1}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Starting the paper
\newpage
\doublespacing

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Introduction section
\noindent
Public universities educate the majority of higher education students in the US, yet have experienced a secular decline in state funding (per student) the last three decades.
This decline has been shown to lead to worse education and later-life outcomes for students \citep{NBERw23736,NBERw27885}, yet it is not clear what mechanisms drive these effects, or how public universities are affected as institutions.
I show that four-year, degree-granting public universities have experienced falling state funding per student, and this affected their composition of faculty.
Falls in public university revenues from the state, instrumented by state-level finance shocks, lead to a fall in the number of tenure-track or tenured professors per student within a university, an increase in the number of lecturers, and overall fall in the number of faculty per student.
US private universities were not exposed to such financial constraints during this same time period, and do not exhibit the substitution away from tenure-track and tenured professors, so that the stagnating state support for public universities has implications for the wider structure of higher education instruction and research in the US.

US universities are widely considered the highest performing in the world, yet there are consequential differences between its universities that operate in the private sector and those established by state governments.
Public universities are subject to numerous state-level administrative laws, and rely on their state governments for funding: an average public university received around \$11,600 per enrolled student in 1990, yet only \$8,300 per enrolled student in 2021.
This fall is driven by stagnating funding provided by state governments, while enrolment among public universities rose by 46\% over the same time period.
At the same time, the number of professors per student at public universities stayed relatively stable at 0.045, yet fell from around 0.04 to around 0.35 professors per student at public universities, driven by a fall of over 20\% in the ratio of associate and full professors per student at public universities.

I use a variant of the \cite{NBERw23736,NBERw27885} shift-share instrument for state funding from shocks to appropriations for the entire state to identify the effect on the faculty composition.
State appropriations for public universities are plausibly endogenous to many outcomes: state governments decide on yearly budgets for their higher education sector, and this process can be influenced by local financial conditions, or even public perceptions of the state's higher education system.
The shift-share instrument interacts reliance on state funding in a base period with the yearly total appropriations per student in the entire state, to exploit both changes in support for higher education and how much each university relies on that support.

Findings show that negative changes in state funding affects faculty composition at public universities, away from tenured and tenure-track professors towards non-tenure track lecturer positions.
A reduction of 10\% in state funding, via a shock to state appropriations, leads to a fall in 1.4\% in the number of assistant professors per student within a university, 1.2\% for full (tenured) professors, and an increas in 4.4\% the number of lecturers per student.
Additionally, I use individual level data on professors at Illinois public universities over 2010-2021 to investigate whether the shocks to state funding affected incumbent professors at public universities.
Incumbent professors were not meaningfully affected, in terms of total salary, promotion rate, or rate of leaving the Illinois public university system, yet lecturers' first year slary decreased by 3.8\% in response to a fall in state funding of 10\%.
These findings imply that changes in faculty composition came about by changes in the composition of hiring new professors at the university level.

\cite{NBERw23736} first study the effect of (changes in) public universities' finances on student outcomes by isolating changes in public universities' expenditures on student instruction, addressing endogeneity of expenditures decisions by instrumenting expenditures with state appropriation shocks and limits on tuition price.
\cite{NBERw23736} find that increases in public university spending (via appropriation shocks) increases enrolment and degree completion among students, but corresponding changes in attendance price (via limits on tuition price) do not have an effect.
\cite{chakrabarti2018effect,NBERw27885} use the same instrument for public university revenues at the state level, combined with rich information on students' individual-level outcomes, to show that increases in state appropriations lead to degree lower completion time and later-life student debt.
\cite{bound2019public} use another variant of the same approach to show that state-level higher education spending cuts induced public universities to shift toward tuition as their primary source of revenue, and shift institutional resources towards students who pay higher tuition rates.
In a similar vein, \cite{bound2007cohort} document variation in per student state funding for higher education resulting from yearly changes in a state's student-age population; they find that lower per student funding lead to lower higher education completion rate among students.

Faculty are a core component of this country's educational and research-innovation university system, yet their composition within universities, and its relationship with higher education finances has so far not been studied.
\cite{brown2014endowment} presents the closest example by studying how university endowments react to negative financial shocks, exhibiting an association between negative endowment shocks and a fall in tenure-system faculty employed at private universities.
\cite{abe2015implications} present a model for university administrator decision-making, and posit that a large component of variation in faculty salaries arises from university politics, and not only economic factors such as outside options.
\cite{johnson2009jep,NBERc13879} both empirically document allocation of faculty between research and instruction, and \cite{hemelt2021math} document differences in instructional cost (per student) between departments, primarily thanks to department differences in class size and faculty salaries.
\cite{turner2014impact} document how universities reacted to the Great Recession of 2008, a severe financial shock to state finances and support for higher education.

The number of professors, both in absolute terms and relative to number of students, may be particularly important for quality of instruction and thus educational outcomes.
\cite{angrist1999using} first causally show that reducing class size (via religious ruling) induces an increase in test scores for school-age children, yet the magnitude of this effect is not as clear in the higher education setting.
\cite{bandiera2010heterogeneous} use a fixed effect model, combined with multiple individual observations in a long panel, to find that UK university students perform worse academically in particularly large classes, and the difference is largest for students at the top of the academic distribution.
For the composition of faculty teaching university classes, \cite{bettinger2010does,figlio2015tenure} find US students have better enrolment and learning outcomes from courses taught by contingent or adjunct (i.e. not tenured or on tenure-track) professors (relative to tenured or tenure-track).
\cite{ehrenberg2005tenured} observe a negative association between utilisation of non-tenured instructors and student graduation rates.
My approach focuses on count of professors per student at universities, so does not directly observe individual class-sizes, yet presents considers this a possible mechanism through which stagnating state support for higher education negatively affects student outcomes \citep{NBERw23736,NBERw27885}.

%\begin{table}[H]
%    \centering
%    \caption{Table to Consider Impact of a Given Paper.}
%    \begin{tabular}{l | c c c c c}
%        & concepts & relationships & models & theories & use of methods \\
%        \midrule
%        
%        change:
%        & concepts & relationships & models & theories & use of methods \\
%        
%        challenge:
%        & concepts & relationships & models & theories & use of methods \\
%        
%        fundamentally alter:
%        & concepts & relationships & models & theories & use of methods \\
%    \end{tabular}
%\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Data section
\section{Data and Institutional Context}
\label{sec:data}

\subsection{Data Description}
The data used in this analysis come from two primary sources: \citet[IPEDS]{ipeds} regarding institutional information on finances and enrolment, and \citet[IBHED]{ibhed} for data on every professor in the Illinois public university system.

IPEDS is a survey of higher educational institutions in the US, and legally requires institutions to participate in order to receive Federal Title IV student aid.\footnote{
    This statement means that IPEDS does not necessarily cover the universe of US higher education institutions, yet in practice every public university and not for-profit four-year institution is represented.
}
Data are consistent between the years 1990 and 2021,\footnote{
    The years 1987-1989 are represented in these data in an incompletion fashion, so I focus on the years 1990 onwards.
    Year refers to the calendar year of the spring term --- i.e. 1987 refers to the academic year that ran August 1986 to July 1987.
}
and provide information on the total revenues a university received from every source (including state governments), enrolment number, plus faculty count and total expenditures on salaries.\footnote{
    I combine the Urban Institute's 2018 compilation of IPEDS data for the years 1990-2017, and manually combine with raw NCES data regarding the year 2018-2021 for all relevant variables.
    Figures for enrolment, faculty counts and salaries come from the raw NCES version of IPEDS for all years to address inconsistencies in the Urban Institute's data formulation for these variables.
}
I restrict analysis to public, four-year, degree-granting institutions, as these institutions adhere to a standardised concept of faculty profile, where tenure and title of appointment (lecturer, assistant professor, etc.) are relatively standardised.
For-profit institutions employ and enrol a negligible share of professors and students respectively, while students at two-year institutions by majority intend to eventually enrol at a four-year institution \citep{mountjoy2022},\footnote{
    Furthermore, \cite{mountjoy2022} documents the effects of expanding two-year higher education access in the US separately for students who would have and those who would not have otherwise have attended college.
    The high rate of eventual enrolment at four-year institutions among two-year students explains the negative impact of increased access to two-year institutions on students who are diverted away from four-year institutions.
}
so that these institutions are not considered.

IPEDS reports the count of professors employed by position, as well as total salary expenditures by position.
This gives a resulting panel data-set, where each row represents a university-year, and includes columns for university finances, plus total count and average salary\footnote{
    Real salary is computed by scaling nominal salary to 2021 dollars by the CPI-U.
} for professors by position (lecturer, assistant, tenured, total).
\autoref{tab:ipeds-summary} presents summary statistics for relevant variables in these data.

\begin{table}[h!]
    \singlespacing
    \centering
    \caption{IPEDS Summary Statistics, Public Universities Panel 1987--2021}
    \makebox[\textwidth][c]{\input{tables/ipeds-summary-fte.tex}}
    \label{tab:ipeds-summary}
\end{table}

IPEDS provides information at the university level, yet lacks information on the distribution of salaries and professor count within a university.\footnote{
    IPEDS provides total paid to salary and total faculty employed per university-year, so that I can investigate average professor salary per university-year with IPEDS data.
    Unfortunately, this measure of professors' salaries is particularly crude in measuring on professors' salaries and other individual-outcomes; summary statistics on IPEDS data do not agree with trends in average professor salary over the sample time period \citep{aau2021survey}.
    Similarly, IPEDS reports of the average salary paid to professors at private institutions also disagree with other sources, so are not considered in this analysis.
}
To investigate the distribution, I integrate individual-level data for every public university professor in the state of Illinois between the years 2010-2021.
IBHED hosts the information;
Public Act 96-0266 (effective 1 January 2010) requires that each university report base salary and benefits all administrators, faculty members, and instructors employed by the college or university.\footnote{
    The universities included are all campuses of the nine Illinois public universities: Chicago State University, Eastern Illinois University, Governors State University, Illinois State University, Northeastern Illinois University, Northern Illinois University, Southern Illinois University  (all five campuses), University of Illinois (all four campuses), Western Illinois University.
}
This publicly available data provides the basis to build a panel of Illinois public university professors 2010-2021; I define a professor as an individual by their first plus last name and university pairing, and link this database to IPEDS regarding finances for their employing institution.

The analysis sample represents 16,932 professors in the year 2010 and 15,352 in the year 2021, with summary statistics presented in \autoref{tab:illinois-summary}.
Additionally, further analysis using the first year of a professor's employment focuses on the subset of professors with observed year of hiring between the years 2011 and 2021, representing 1,778 professors in 2011, and 9,099 in 2021.\footnote{
    Summary statistics for this group, which over-represents lecturers compared to assistant and full professors, can be seen in \autoref{tab:illinois-summary-rolling}.
    The reasons for restricting to the sample with identified year of hire in the range 2011-2021 is explained in \autoref{sec:iv-model-indiv}.
}

\begin{table}[h!]
    \singlespacing
    \centering
    \caption{IBHED Summary Statistics, Professor Panel 2010--2021.}
    \makebox[\textwidth][c]{\input{tables/illinois-summary.tex}}
    \label{tab:illinois-summary}
\end{table}

\subsection{Trends in Finances, Enrolment, and Faculty Composition}
\label{sec:trends}
A state government plans an annual budget a couple of years ahead of the fiscal year, and the legislature chooses to approve or reject a budget request put forth by the governor's office.\footnote{
    \cite{NBERw23736} present a full discussion of the state appropriation decision-making process, drawing on administrative records originally analysed by \cite{parmley2009state}.}
US state governments, by majority, are legally obligated to run a balanced-budget, so that yearly variation in tax revenues (caused by changing economic conditions or otherwise) necessarily affect state expenditures.
This process leads to yearly variation in state appropriations not seen in other revenues sources (such as federal government appropriations), since US states differ in their support for higher education, and are subject to fiscal constraints.
State appropriations to higher education are a particularly attractive area of state spending to absorb such shocks to state finances \citep{delaney2011state}.\footnote{
    \cite{delaney2011state} fully describes the financial environment of state expenditures, and what makes spending on higher education an attractive area for state governments to expand funding during years of higher tax revenues, and retract funding in leaner years.
    An analysis of state expenditures for the years 1980-2004 (overlapping with the sample for this analysis) provides solid evidence for these trends.
}
Additionally, there is yearly variation in the number of higher education students in each state \citep{turner2014impact}, so that per-student state appropriations vary on multiple dimensions.

\autoref{fig:funding} exhibits the trends for the mean public university for the years 1990-2021.
We see a rise in total revenues received by public universities (from all sources), and a notable increase in mean tuition revenues from \$48 million per year-university to \$150 million per year-university.
At the same time, total state appropriations stagnated at around \$100 million per year-university for 1990-2008, falling around 2008 and have not recovered ever since.
While public universities experienced a stagnation in state support, private universities were not exposed to the same constraints, receiving \$37,000 per student in 1990 and \$49,000 in 2021, experiencing no corresponding decline in any specific component.

\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{Mean Total Revenues among Public Universities, by Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Total, millions \$ 2021 CPI-U.}
        \includegraphics[width=\textwidth]{figures/mean-funding-total.png}
        \label{fig:mean-funding-total}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Per Enrolled Student, \$ 2021 CPI-U.}
        \includegraphics[width=\textwidth]{figures/mean-funding-fte.png}
        \label{fig:mean-funding-fte}
    \end{subfigure}
    \label{fig:funding}
\end{figure}

\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{Total Student enrolment, by University Sector, and Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Total.}
        \includegraphics[width=\textwidth]{figures/enrollment-total.png}
        \label{fig:enrollment-total}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Mean, University Level.}
        \includegraphics[width=\textwidth]{figures/enrollment-mean.png}
        \label{fig:enrollment-mean}
    \end{subfigure}
    \label{fig:enrolment}
\end{figure}

At the same time, student enrolment at public universities rose precipitously.
6.2 million students were enrolled in public universities in 1990, and this number rose by 47\% to 9.1 million, with most of the increase occurring over the years 2000-2021.
\autoref{fig:enrolment} shows that total enrolment at private universities has also risen over the same time period, but not as drastic in either relative or absolute terms; the mean private university grew from 9,800 students in 1990 to 11,800 in 2021.
This means that revenues per student have stagnated across all measures (seen in \autoref{fig:mean-funding-fte}), and particularly fallen from a mean of \$11,000 per student in 1990 to less than \$8,000 per student in 2021.

\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{Total Professor Count per Student, by University Sector, Professor Appointment, and Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Lecturer}
        \includegraphics[width=\textwidth]{figures/lecturer-fte-perprof.png}
        \label{fig:lecturer-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Assistant.}
        \includegraphics[width=\textwidth]{figures/assistant-fte-perprof.png}
        \label{fig:assistant-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Full.}
        \includegraphics[width=\textwidth]{figures/full-fte-perprof.png}
        \label{fig:full-fte-perprof}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{All.}
        \includegraphics[width=\textwidth]{figures/all-fte-perprof.png}
        \label{fig:all-fte-perprof}
    \end{subfigure}
    \label{fig:fte-perprof}
\end{figure}

\autoref{fig:fte-perprof} documents the divergence in faculty composition (per student) between the average private and public university 1990-2021.
Private universities start with  a higher baseline of around 4.5 professors per 100 students, and exhibited yearly variation of less than 0.5 professor per hundred students over the thirty years.
Public universities start with 3.9 professors per hundred students, and this number falls to around 3.5 primarily in the 2008-2011 time period.
We see a similar difference in baseline, and fall for the years 2008-2011 for assistant professors.
Private and public universities have similar numbers of associated and full professors before the year 2000, yet this number has fallen by over 20\% in the next 20 years only for public universities: in 2021 the mean public university has 6 fewer full professors per hundred students than the mean private university.
Over the same time period, we see the rise in use of non-tenure track instructor positions (referred to as lecturers from here on), who were employed at similar rates in both sectors in 1990 yet have been utilised by public universities at a higher rate since.

\subsection{Trends in Illinois}
\label{sec:trends-illinois}

The state of Illinois experienced the same stagnation in state support for higher education that the rest of the nation has experienced 1990-2021.
\autoref{fig:illinois-funding} shows the trends for revenues among the seven Illinois public university campuses, including the falling share of state appropriations and substitution towards tuition revenue.

\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{Mean Revenues among Illinois Public Universities, by Year.}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Total, millions \$ 2021 CPI-U.}
        \includegraphics[width=\textwidth]{figures/illinois-funding-total.png}
        \label{fig:illinois-funding-total}
    \end{subfigure}
    \begin{subfigure}[b]{0.495\textwidth}
        \centering
        \caption{Per Enrolled Student, \$ 2021 CPI-U.}
        \includegraphics[width=\textwidth]{figures/illinois-funding-fte.png}
        \label{fig:illinois-funding-fte}
    \end{subfigure}
    \label{fig:illinois-funding}
\end{figure}

The decade of 2010-2021 has also seen yearly variation in its revenues, particularly around 2016.
In the calendar year 2015 partisan disagreements between the democratic legislature and republican governor led to the 2016 fiscal year starting with no state budget.
State agencies, and higher education institutions, employed accounting techniques to continue operating without any resources provided by the state government.\footnote{
    Fiscal year 2016 refers to June 205 to June 2016, as is the same for the academic year definition in this analysis. 
    K-12 schools were unaffected thanks to a separate budget agreement allowing non-higher education to operate as before.
}
While state institutions were able to stay open, there were drastic revenue and spending cuts in response to the budget impasse, as it continued through fiscal year 2017, and ended with a new budget restoring revenues to state agencies and universities for the 2018 fiscal year.
This variation in public university revenues via the state appropriations channel, stemming entirely from political disagreements --- not from state decisions regarding higher education and its finances \citep{young2020squandered} ---  mean that Illinois public universities exhibit sizable changes in their state appropriations over 2010-2021, of similar order to those for the rest of the country over 1990-2021.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Section on possible outcomes
\subsection{University Responses to Financial Shocks}
\label{sec:responses}

While state support for public higher education has stagnated at the same time as education costs rose, there are multiple possible ways that a university can respond.
\cite{NBERw23736} established that corresponding rises in tuition did not offset the falling state support, so that university spending (per student) fell in response to these persistent, negative state funding shocks.
There are multiple ways that these changes in finances may affect the faculty composition at public universities.

Universities every year hire a number of new professors, either to expand their departments or to replace leaving professors.
The number of professors hired is usually highest among non-tenured adjunct faculty, as these instructors are by majority hired on short-term contingent contracts, so that they are less costly to hire (or to not reinstate) in response to yearly changes in teaching needs.
Secondly, tenure-track assistant professors are hired in most years at public universities to replace leaving assistant professors, and are most often hired with a four to six year contract and agreement to be considered for tenure at the end of this term.
Lastly, tenured professors are those that have undergone review and have successfully secured a full-time appointment at their university with no expiration date; tenured professors are sometimes hired from outside a university, though this position has the lowest hiring rate among most universities.
It is common for universities to restrict their faculty hiring in the face of financial shocks.
Since the yearly hiring rate is highest among lecturers and assistant professors, financial shocks may lower the number of lecturer and assistant professors at public universities the most, if faculty hiring is affected by falls in state funding.

Additionally, professors' salaries may be disrupted by the stagnation in public university finances.
When the university has a lower budget for its yearly hiring, it may respond by lowering the salary they offer to new hires.
This possible effect may not be the same across each position of professor; tenured faculty are often hired away from another university, so that new tenured professor hires could be less likely to accept a lower offer from a public university, and we would see that new tenured hires are not affected salary-wise.\footnote{
    \cite{blackaby2005} describes the academic outside options among economist faculty, documenting how they differ for academic position and gender.
}
Yet salary for all the professors, not just new hires, may also be affected: multiple universities passed a university-wide pay-cut for their faculty in response to state budget cuts around the 2008 recession, for example.\footnote{
    Indeed, Cornell University implemented both hiring freezes and even a nominal salary reduction for professors, in anticipation of a financial shock in early 2021.
    The salary cut was not permanent, as the oncoming financial shock turned out to not be as serious as projected, so that the salary cut was returned to professors in the year 2021.
}
Faculty are not paid the same by position, so that their are financial consequences for faculty being promoted between positions within a university.
In response to tightening fiscal restraints, universities may be less likely to grant tenure to assistant professors, or to promote tenured professors from associate to full professor either.

It is not immediately clear which effect will dominate, and which position of professor will be effected the most by changes in their university's finances.
Yet, there is one empirical fact worth noting: a mean assistant or tenured professor earns more than double that of an adjunct lecturer (\autoref{tab:illinois-summary}), and has more long term salary obligations than a lecturer too.
If a public university's primary obligation is to teach, and they must fulfil this objective with fewer and fewer resources, then they may substitute away from tenure-track and tenured professor towards contingent lecturers to achieve this obligation.\footnote{
    That is substituted away from tenure-track and tenured professors according to the salary ratio, and the relative productivity between these two positions.
    It is not possible to estimate the relative productivity in this paper, so I leave the rate of substitution between the two positions to further research.
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Empirical section
\section{Empirical Framework}
\label{sec:empirics}

A professor is employed by a university, so it stands to reason that when public universities experience a secular fall in one of their revenue soruces revenue --- analogous to a demand shock in the private sector --- that the composition of their employees will be affected.

Na\"ively we can analyse the relationship between professor-outcomes $Y_{i,t}$ at university $i$ in year $t$ as a result of state funding per student $X_{i,t}$.
Outcomes include count of professors per student, and average salary, within each university-year.
This model gives $\beta$ the association with these outcomes and state funding per student,\footnote{
    Note that dividing by student count also implicitly controls for the size of the university, so that this model accounts for yearly variation in professor count and university revenues arising from growth in a university.
}
including fixed effects to control for effects specific to the university and year.
Log$(.)$ transforming the variables improves the interpretability of $\beta$ as an elasticity for professor count per student with respect to state funding per student.
\begin{equation}
    \label{eqn:naivereg}
    Y_{i,t} = \alpha_i + \gamma_t + \beta X_{i,t} + \epsilon_{i,t}
\end{equation}

Yet a university's finances are not exogenous to state decisions on the yearly level for support of higher education, or exogenous to internal institutional decisions.
Instead, the state government and university administration undertake a complex process of allotting resources across multiple different priorities, including instruction, research, or between departments.
Importantly for this analysis, revenues received from an institution's state government provide opportunity to address this endogeneity.

\subsection{State Appropriation Shocks}
\label{sec:approp-shocks}

\cite{NBERw23736,chakrabarti2018effect,NBERw27885} address endogeneity in public university finances by exploiting a shift-share instrument for changes in state-level funding interacted with university reliance on state funding in a base period.
The system exploits the fact that institutions who rely on state appropriations more will be affected by state appropriation shocks.
\begin{align}
    \label{eqn:public-instrument}
    Z_{i,t} &\coloneqq \log \left[
    \left( \frac{\text{Total State Funding}_{s(i),t}}{\text{Student Population}_{s(i),t}} \right)
    \sum_{\tau = 0}^{3} \frac 14
    \left( \frac{\text{State Funding}_{i,1990 + \tau}}{\text{Total Revenues}_{i,1990 + \tau}} \right) \right]
\end{align}

$Z_{i,t}$ is the instrument for (log) state appropriations for institution $i$ in year $t$, interacting the average funding for university $i$ in state $s(i)$ with reliance on state funding relative to total revenues, averaged across the base years 1990--1993.\footnote{
    1990--1993 are defined as data for public university finance data are most comparable (i.e. without many missing values) beginning in 1990.
    \cite{NBERw23736} use the single year 1990 as the base year, though I use the four years to ameliorate missing values in the single year of 1990.
    Results are similar in either specification.
}
\cite{NBERw27885} notes the tendency for public universities to respond to state funding cuts by increasing reliance on tuition, where \cite{NBERw23736} specifically instruments for tuition revenues with collected information on legislative tuition price controls.\footnote{
    \label{foot:control}
    It may be argued that tuition revenues are confounder between the causal effect of changes in state funding on a university's total revenues, so that this analysis focuses on state support for higher education (no total revenue) as a result.
    This follows the \cite{NBERw27885} formulation.
    On the other hand, rises in tuition revenues (per student) may arise as result of tuition hikes thanks declining state support, which would mean controlling for tuition would constitute a bad control.
    Estimates including tuition revenues (per student) as a control in the second stage of the IV estimates produces results of very similar magnitude and direction, and so are omitted.
}
This analysis focuses on one important source of revenues for public universities: state appropriations.

The first-stage is then as follows, including institution and year fixed effects.
\begin{equation}
    \label{eqn:firststage}
    X_{i,t} = \eta_i + \zeta_t + \delta Z_{i,t} + \epsilon_{i,t}
\end{equation}
We note the conditions for exogeneity in the instrument (following the discussion presented by \citealt{NBERw27885}).
The instrument is exogenous if state policy decisions for funding of public universities are uncorrelated with unobserved institutional changes of any specific college or university in the state \citep{borusyak2022quasi}.
This assumption is plausible given that the majority of states have multiple (i.e. more than five) public universities, without any single university campus receiving the majority of state funding within any single state.
Secondly, addressing endogeneity in university revenues by the shift-share identification strategy requires exogeneity in either the base-line share or shift component of the instrument.
In this case, we satisfy the second: universities' institutional-level decisions are not correlated with contemporaneous or upcoming shocks to state appropriations.\footnote{
    It would be plausible to consider the case that universities make institutional-level decisions in a consistently different manner to those 
    with differing reliance on state appropriations in 1990, so that exogeneity by the base-line share is not plausible here.
}

\begin{table}[h!]
    \singlespacing
    \centering
    \caption{First Stage Estimates, for University Revenues by Appropriation Shock.}
    \makebox[\textwidth][c]{\input{tables/firststage-reg-fte.tex}}
    \label{tab:firststage-reg}
    \begin{flushleft}
        \footnotesize
        \textbf{Note}: Standard errors are clustered at the state-year level.
    \end{flushleft}
\end{table}

\autoref{tab:firststage-reg} presents results of the first-stage regression, separately with and without a control for per student tuition revenue and year plus institution fixed effects.\footnote{
    Representations for frequentist significance levels (i.e. 10, 5, 1\% etc.) are omitted here, and in all following tables.
}
We note columns (1) and (2) estimate that a shock to state appropriations per student in the state of 10\% is associated with around 9.8\% change in state funding per student at the university; the instrument is strong, and we note similarity in estimates with and without inclusion of fixed effects.
Columns (3) and (4) include the tuition revenue control (explained in \autoref{foot:control}) to exhibit estimates with the inclusion of this possible collider or bad control.
Column (3) shows similar estimates to columns (1), (2) thanks to inclusion of fixed effects, so that the fixed effects was effective in soaking variation in per-student tuition revenues at the institution-year level.
Column (1) represents the estimates for \autoref{eqn:firststage} with fixed effects, omitting the tuition revenue control, and is the preferred form that I proceed with in the following.

\begin{figure}[h!]
    \centering
    \singlespacing
    \caption{Local Projection Estimates for First-Stage \autoref{eqn:firststage}.}
    \includegraphics[width=0.6\textwidth]{figures/firststage-lp.png}
    \label{fig:firststage-lp}
\end{figure}

\autoref{fig:firststage-lp} presents local projection estimates for the staying-power of the state appropriation shock to university revenues.
We see the time 0 estimate coincides with that of Column (1) in \autoref{tab:firststage-reg}, the contemporaneous effect of the shock revenues.
We see a decaying, yet real and positive, effect of the shock on revenues for the next 10 years, illustrating that the state appropriation shock is a strong, if fading, instrument for state funding per student and justifying its use in local projection IV estimation.

These results, together with the case for exogeneity, show the appropriations shock instrument strongly predicts university revenues in the first-stage estimation.

\subsection{Instrumental Variables Model, University-Level}
\label{sec:iv-model-uni}

The primary empirical model combines the instrumental variable for state appropriation shocks with the empirical model for the effects of state funding at the university-level --- i.e. parameter $\beta$ in the following.\footnote{
    It is important to note the treatment effect isolated here; the instrumental variables approach identifies the local average treatment effects (LATE) specific to the instrument.
    So we interpret this treatment effect as a university's response in employment count and average salaries to state funding changes, changes specific to state appropriation shocks, among the complier group --- i.e. universities with any exposure to shocks, assuming no universities increase employment or salaries in response to negative revenue shocks.
    \cite{mogstad2021causal} provide a full discussion of the LATE with instruments and multiple monotonicity conditions.
}
\begin{eqnarray}
    \label{eqn:secondstage1}
    X_{i,t} &=& \eta_i + \zeta_t + \delta Z_{i,t} + \epsilon_{i,t} \\
    \label{eqn:secondstage2}
    Y_{i,t} &=& \alpha_i + \gamma_t + \beta \widehat X_{i,t} + \varepsilon_{i,t}
\end{eqnarray}
I estimate the system by two stage least squares, including institution and year fixed effects, and investigate outcomes at the university-level to analyse effects on the university as a result of changes in revenues.
Additionally, I estimate the model via local projections \citep{jorda2005,miller2022} to investigate whether the effects of revenues on faculty composition linger for multiple years after the original appropriation shock.
Regarding outcomes, I focus on the composition of the professors employed at the university by analysing (log) count per student, and average salaries paid to, professors employed by the university.

\begin{figure}[H]
    \centering
    \singlespacing
    \caption{Instrument Variables Model for University Finances and Faculty Composition.}
    \begin{tikzpicture}
        \node[state] (instrument) at (0,0) {$Z$};
        \node at (-2.6,0.5){State};
        \node at (-1.9,0){appropriation};
        \node at (-2.6,-0.5){shock};
        \node[state] (endogenous) [right=of instrument] {$X$};
        \node at (2,1.5) {University};
        \node at (2,1) {revenues};
        \node[state] (outcome) [right=of endogenous] {$Y$};
        \node at (5.2,0.25) {Faculty};
        \node at (5.6,-0.25) {Composition};
        % Relevance condition
        \path (instrument) edge (endogenous);
        \path (endogenous) edge (outcome);
        % Exclusion Restriction
        \path (instrument) edge[bend right=60] (outcome);
        \node[color=red] at (2,-1.2) {X};
    \end{tikzpicture}
    \label{fig:SCM-ivmodel}
\end{figure}

>>>>

@@

\def\equationautorefname~#1\null{Equation~(#1)\null}                       % Fix autoref for equations

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Title page
% Author
\author{Senan Hogan-Hennessy\thanks{
    Economics Department, Cornell University.}
}
% Title
\title{Changes in Faculty Composition and Stagnating State Support for Higher Education
    }
\date{\today\thanks{
    I thank Professor Evan Riehl for guidance on this project, Michael Lovenheim, Ronald Ehrenberg, Douglas Miller, Francine Blau, and Ryan Dycus for various comments on the project's direction, as well as seminar participants at Cornell University (Autumn 2022) for helpful discussion.
    This project's Github repository, with materials and (publicly available) data for replication, is available at 
    \url{https://github.com/shoganhennessy/state-faculty-composition}.
    Any comments or suggestions may be sent to me at \href{mailto:seh325@cornell.edu}{\nolinkurl{seh325@cornell.edu}}, or raised as an issue on the Github project.
    }
}
\rhead{\today}
\lhead{Faculty Composition and State Support for Higher Education}
% Begin
\begin{document}
\clearpage
\maketitle
\thispagestyle{empty}
% Abstract
\begin{abstract}
    \input{sections/00-abstract.tex}
\end{abstract}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Starting the paper
\newpage
\setcounter{page}{1}
\doublespacing
% Introduction section
\input{sections/01-introduction.tex}
% Data section
\input{sections/02-data.tex}
% Conceptual Framework
\input{sections/03-concepts.tex}
% Empirical methods
\input{sections/04-empirical.tex}
% Empirical results
\input{sections/05-results.tex}
% Dicussion
\input{sections/06-discussion.tex}
% Conclusion
\input{sections/07-conclusion.tex}
% Bibliography
\bibliographystyle{agsm}
\bibliography{sections/08-bibliography.bib}
% Appendix
\input{sections/09-appendix.tex}
\end{document}
